Reconstructing contact network parameters from viral phylogenies

### 2016-07-14 doc/paper/paper.tex
Finished reviewer things finally ugh

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Wed Jul  6 16:31:52 2016
\begin{table}[ht]
\centering
\begin{tabular}{lllll}
  \hline
Dependent variable & Independent variable & Estimate & Standard error & $p$-value \\ 
  \hline
$\alpha$ & (Intercept) & $2$ & $0.6$ & $0.01$ \\ 
   & $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
   & $I$ & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
   & $m$ & $0.5$ & $0.2$ & $0.01$ \\ 
  $I$ & (Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
   & $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
   & $I$ & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
   & $m$ & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
   \hline
\end{tabular}
\caption[Parameters of fitted GLMs relating error in estimated $\alpha$ and $I$ to true values of BA parameters.]{Parameters of fitted GLMs relating error in estimated $\alpha$ and $I$ to true values of BA parameters. GLMs ere fitted with the Gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glm}
\end{table}

@@

\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
of each of the BA parameters on the same simulated transmission tree. The
estimates of $m$ were apparently unaffected by marginalizing out the other
parameters, corroborating the previous experiments' findings that $m$ is not an
identifiable parameter from scaled tree shapes. Compared to allowing all
parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
    41\%,
    59\%, and
    46\%
when all other parameters were fixed. \Cref{fig:better} shows the impact of
increasing the number of particles, simulated datasets, and
$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
number of iterations until the stopping condition was reached was 81 with the
basic settings and 124 with the higher settings. The results of the two
simulations were very similar for all parameters.}

@@

\add{Posterior mean point estimates for the preferential attachment power $\alpha$
were all sub-linear, ranging from 
    0.27
(mixed/Spain) to
    0.81
(IDU/Estonia). When aggregated by risk group, the average estimates were
    0.77
for IDU,
    0.41
for primarily heterosexual risk, and
    0.37
for MSM. 95\% HPD intervals were very wide for most datasets, often
encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}).}

@@

\add{For all but the HET/Botswana data, the posterior mean estimates for
$I$ were between
    410
(MSM/Taiwan) and
    1391
(MSM/Shanghai). The HET/Botswana data had a much higher estimated $I$ value
    (5432)
than the other datasets, with a very wide 95\% HPD interval covering almost the
entire prior region (\cref{fig:abchpd}). There was no significant correlation
between the number of sequences in the tree and the estimated $I$ value
(Spearman correlation, 
    $p =0.8$),
indicating that the higher estimates were not simply due to increased sampling
density. When both \textit{gag} and \textit{env} sequences were analyzed, the
estimates from the \textit{env} data were higher (HET/Uganda,
    939 for \textit{gag} vs.
    1615 for \textit{env};
HET/Malawi,
    724 for \textit{gag} vs.
    845 for \textit{env}).}

@@

\add{The posterior means of $m$ were equal to one for 
  zero
of the datasets analyzed. The widths of the 95\% HPD intervals varied from 0
(all the mass on the estimated value) to 5 (the entire prior region). Estimates
of $N$ were  mostly uninformative, with very similar estimates for all datasets
(mean
    6188,
range
    5881 - 6882).
This was similar to the pattern observed for the synthetic data, where the
posterior mean always fell around the upper two-thirds mark of the range
(\cref{fig:abcpt2}).}

\add{When the value $m = 1$ was disallowed by the prior, the separation in
$\alpha$ between the IDU datasets and the others became more striking
(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
or above
    1.
The estimate for the MSM/Beijing data was slightly lower
    (0.85) and
the estimates for the seven remaining non-IDU datasets were bounded above by
    0.58. 
The values of $I$ were fairly robust to the choice of prior (compare
\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
narrower (average width
    2192
for $m \geq 1$ and
    1874
for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
took on the value 3 with this prior, with the HPD intervals spanning the entire
prior region. This is very similar to the results observed for $m$ on simulated
data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
data with this prior. The results for $N$ did not change appreciably between
the two choices of prior.}

@@

\add{For all datasets we examined, the posterior mean estimates for $\alpha$
were sub-linear, ranging from 
    0.27
to 
    0.81.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated preferential attachment model
incorporating heterogeneous node behaviour. When used to analyze
population-level longitudinal partner count data, they found $\alpha$ values
ranging from 0.26 to 0.62 depending on the gender and time period considered.}

@@

\add{The true HIV prevalence in a population can be difficult to estimate for
several reasons. HIV-infected individuals may be asymptomatic for months or
years, possibly delaying their awareness of their status. In many contexts, the
risk factors for acquisition of HIV are illegal or stigmatized, which may
represent a barrier to testing, treatment, and/or disclosure of status.
Our simulation study showed that $I$ is weakly identifiable from tree shapes,
however, the estimates of $I$ obtained with \textit{netabc} were upwardly
biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
that the identifiability of $I$ decreases with the number of sampled tips
(\cref{fig:Icrossv}); in real world studies, the proportion of infected
individuals sampled is usually low. The estimated $I$ values for the HIV
datasets ranged from
    410
    (IDU/Estonia)
to
    5432
    (HET/Botswana).
The estimates of $I$ were less robust to the HIV gene analyzed
(\cref{fig:genes}) and the phylogenetic reconstruction method used
(\cref{fig:methods}).}

>>>>

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Thu Jul 14 15:34:47 2016
\begin{table}[ht]
\centering
\begin{tabular}{lllll}
  \hline
Dependent variable & Independent variable & Estimate & Standard error & $p$-value \\ 
  \hline
$\alpha$ & (Intercept) & $2$ & $0.6$ & $0.01$ \\ 
   & $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
   & $I$ & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
   & $m$ & $0.5$ & $0.2$ & $0.01$ \\ 
  $I$ & (Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
   & $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
   & $I$ & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
   & $m$ & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
   \hline
\end{tabular}
\caption[Parameters of fitted GLMs relating error in estimated $\alpha$ and $I$ to true values of BA parameters.]{Parameters of fitted GLMs relating error in estimated $\alpha$ and $I$ to true values of BA parameters. GLMs ere fitted with the Gaussian distribution and inverse link function. Coefficients are interpretable as additive effects on the inverse of the mean error.} 
\label{tab:glm}
\end{table}

@@

\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
of each of the BA parameters on the same simulated transmission tree. The
estimates of $m$ were apparently unaffected by marginalizing out the other
parameters, corroborating the previous experiments' findings that $m$ is not an
identifiable parameter from scaled tree shapes. Compared to allowing all
parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
    41\%,
    59\%, and
    46\%
when all other parameters were fixed. \Cref{fig:better} shows the impact of
increasing the number of particles, simulated datasets, and
$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
number of iterations until the stopping condition was reached was 81 with the
basic settings and 124 with the higher settings. The results of the two
simulations were similar, but surprisingly, the results with higher SMC
settings were slightly worse (by
    10\%,
    8\%, and
    11\%
for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
$I$ was closer to the true value of 2000 with the improved settings
    (2338 - 
     3423, vs.
     2810 - 
     3767 with basic settings).
The estimate of $m$, 3 in both cases, was unaffected by the settings.}

@@

\add{Posterior mean point estimates for the preferential attachment power $\alpha$
were all sub-linear, ranging from 
    0.27
(mixed/Spain) to
    0.83
(IDU/Estonia). When aggregated by risk group, the average estimates were
    0.78
for IDU,
    0.41
for primarily heterosexual risk, and
    0.37
for MSM. 95\% HPD intervals were very wide for most datasets, often
encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}). As shown
in \cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
analyzed.}

@@

\add{For all but the HET/Botswana data, the posterior mean estimates for
$I$ were between
    373
(IDU/Estonia) and
    1391
(MSM/Shanghai). The HET/Botswana data had a much higher estimated $I$ value
    (5432)
than the other datasets, with a very wide 95\% HPD interval covering almost the
entire prior region (\cref{fig:abchpd}). There was no significant correlation
between the number of sequences in the tree and the estimated $I$ value
(Spearman correlation, 
    $p =0.9$),
indicating that the higher estimates were not simply due to increased sampling
density. When both \textit{gag} and \textit{env} sequences were analyzed, the
estimates from the \textit{env} data were higher (HET/Uganda,
    939 for \textit{gag} vs.
    1615 for \textit{env};
HET/Malawi,
    724 for \textit{gag} vs.
    845 for \textit{env}).}

@@

\add{The posterior means of $m$ were equal to one for 
  zero
of the datasets analyzed. The widths of the 95\% HPD intervals varied from 0
(all the mass on the estimated value) to 5 (the entire prior region). Estimates
of $N$ were  mostly uninformative, with very similar estimates for all datasets
(mean
    6202,
range
    5881 - 6882).
This was similar to the pattern observed for the synthetic data, where the
posterior mean always fell around the upper two-thirds mark of the range
(\cref{fig:abcpt2}).}

\add{When the value $m = 1$ was disallowed by the prior, the separation in
$\alpha$ between the IDU datasets and the others became more striking
(\cref{fig:abchpdm2}). Both IDU datasets had estimated $\alpha$ values at or
above
    1.
The estimate for the MSM/Beijing data was slightly lower
    (0.85) and
the estimates for the seven remaining non-IDU datasets were bounded above by
    0.58. 
The values of $I$ were fairly robust to the choice of prior (compare
\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
narrower (average width
    2159
for $m \geq 1$ and
    1874
for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
took on the value 3 with this prior, with the HPD intervals spanning the entire
prior region. This is very similar to the results observed for $m$ on simulated
data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
data with this prior. The results for $N$ did not change appreciably between
the two choices of prior.}

\add{For the two datasets we reanalyzed using
RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
was relatively robust to the choice of method (\cref{fig:methods}, posterior
means 
    0.48
vs. 
    0.48
for mixed/Spain and 
    1.02
vs.
    1.12
for IDU/Estonia). However, the estimates of $I$ were about twice as high when
RAxML was used instead of FastTree to reconstruct the trees 
    (228 
     vs. 437 for IDU/Estonia, 
    816
    vs. 1949 for mixed/Spain). 
\Cref{fig:boot} shows estimates obtained for five bootstrap replicate
alignments for each of these two datasets. For the mixed/Spain data, the
estimated posterior mean [range of bootstrap posterior means] was
    0.48
    [0.54 -
     0.67]
for $\alpha$, 
    816
    [403 -
     886] for $I$,
    2.76
    [2.61 -
     3.24] for $m$, and
    6639
    [6652 -
     7245] for $N$. 
For the IDU/Estonia data, values were
    1.02
    [0.78 -
     1.07]
for $\alpha$, 
    228
    [313 -
     741] for $I$,
    3.11
    [3.41 -
     3.46] for $m$, and
    6803
    [5941 -
    6913] for $N$.}

@@

\add{For all datasets we examined, the posterior mean estimates for $\alpha$
were sub-linear, ranging from 
    0.27
to 
    0.83.
The sub-linearity is consistent with the results of
\textcite{de2007preferential}, who developed a statistical inference method to
estimate the parameters of a more sophisticated preferential attachment model
incorporating heterogeneous node behaviour. When used to analyze
population-level longitudinal partner count data, they found $\alpha$ values
ranging from 0.26 to 0.62 depending on the gender and time period considered.}

@@

\add{The true HIV prevalence in a population can be difficult to estimate for
several reasons. HIV-infected individuals may be asymptomatic for months or
years, possibly delaying their awareness of their status. In many contexts, the
risk factors for acquisition of HIV are illegal or stigmatized, which may
represent a barrier to testing, treatment, and/or disclosure of status.
Our simulation study showed that $I$ is weakly identifiable from tree shapes,
however, the estimates of $I$ obtained with \textit{netabc} were upwardly
biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
that the identifiability of $I$ decreases with the number of sampled tips
(\cref{fig:Icrossv}); in real world studies, the proportion of infected
individuals sampled is usually low. The estimated $I$ values for the HIV
datasets ranged from
    373
    (IDU/Estonia)
to
    5432
(HET/Botswana).}
